DeepSphere: a graph-based spherical CNN

### 2019-09-25 main.tex
many small edits to experiments

\todo{
\begin{itemize}
	\item shall we really name it? Or simply reference the general idea of a graph-based spherical CNN.
	\item adjacency matrix: $A$ instead of $W$?
	\item anonymize DeepSphere citation
% 	\item climate spherical figure: inputs with TC and AR labels superimposed [Frédérick]
	\item citet for inline, citep for parenthesis
	\item More details on related work? Not sure we have the space. If yes, detailed comparison with related work after method.
	\item PSD of climate data: how does it compare to cosmo and 3D objects?
	\item align the spheres in figure 1
	\item mention that \citep{esteves2018sphericalcnn} and \citep{cohen2018sphericalcnn} computed equivariance error
	\item check footnotes
	\item climate data on spectrum plot
	\item result tables: only what supports the main message in the text, everything we have in the appendix
	\item Unify vocabulary. Sampling, discretization, or pixelization of the sphere? Signal, maps, functions. Irregular, uneven. Coarsening, down-sampling.
	\item check English vs American English
	\item unify enumerations. I like (i), (ii), etc.
	\item notation: should we use indices from 1 to n instead of 0 to n-1 for compactness?
	\item add \citet{bruna2013gnn} in equivariance? They were the first to do a spherical CNN with graphs. Graph is built from a uniform random sampling of the sphere.
	\item figure placements
	\item do the experiments on the icosahedral grid with the pooling as defined in the paper
	\item put all the code (learning experiments from Frédérick and equivariance experiments from Martino) in a common reproducible git repository to be linked from the paper
	\item update pygsp and merge sphere branch
	\item check that NN architectures and hyper-params are correctly reported
	\item keep NN architecture and hyper-params in the main text? Move the paragraph to appendix if not enough space.
	\item shall we still site adam and batch norm?
	\item check results on ModelNet40
	\item Check that HEALPix is introduced in the introduction
\end{itemize}
}
\newpage

@@

\paragraph{Architecture.}
All our NNs are fully convolutional, and employ a global average pooling (GAP) for rotation invariant tasks.
Graph convolutional layers are always followed by batch normalization \citep{ioffe2015batchnorm} and ReLU activation, except in the last layer.
Note that batch normalization and activation act on the elements of $\b{f}$ independently, and hence don't depend on the domain of $f$.
%We denote graph convolutional layers with $C$ output channels as $GC_C$ and fully connected layers with $C$ output neurons as $FC_C$.
%Those linear layers are always followed by batch normalization and ReLU activation, except in the last layer.
%softmax normalization as $SM$
%Pooling and unpooling by a factor $F$ are denoted as $P_F$ and $P^{-1}_F$, global average pooling as $GAP$.

@@

The recognition of 3D shapes is a rotation invariant task: rotating an object doesn't change its nature.
While 3D shapes are usually represented as meshes or point clouds, representing them as spherical maps (\figref{depthmap}) naturally allows a rotation invariant treatment.
LIDAR actually acquires depth maps, which are a posteriori represented as point clouds.
\todo{move to intro?}

@@

The equiangular graph is built with \eqnref{weights:khasanova} and $k = 4$ neighbors \citep[following][]{khasanova2017sphericalcnn}.
\todo{The HEALPix graph is built with \eqnref{weights:belkin}, $k = ?$ and the corresponding optimal kernel width $t = ?$.}
\todo{Following Esteves or Cohen?}
The NN is made of $5$ graph convolutional layers, each followed by a \todo{max?} pooling layer down-sampling by $4$.
A GAP and a fully connected layer with softmax follow.
The Chebyshev polynomials are all of order $K=4$ and the number of channels per layer is $16, 32, 64, 128, 256$, respectively.
Following \citet{esteves2018sphericalcnn}, the cross-entropy plus a triplet loss is optimized with Adam \citep{kingma2014adam} for 30 epochs on the dataset augmented by 3 random translations.
The learning rate is $5 \cdot 10^{-2}$ and the batch size is 32.

@@

Results are shown in \tabref{shrec17}.
DeepSphere achieves the same performance as \citet{cohen2018sphericalcnn} and \citet{esteves2018sphericalcnn} at a much lower cost, suggesting that anisotropic filters are an unnecessary price to pay.
% x times faster
As the information in those spherical maps resides in the low frequencies (\figref{spectrum}), reducing the equivariance error didn't translate into improved performance.
For the same reason, using the more uniform HEALPix discretization or lowering the resolution down to $N_{side} = 8$ didn't impact performance.
\todo{be sure that it fits with the cosmo conclusion}
\todo{The same conclusions are drawn from ModelNet 40 (see appendix ??).}

@@

Graphs are built with \eqnref{weights:belkin}, $k = 8, 20, 40$ and the corresponding optimal kernel widths $t = ?, ?, ?$.
Following \citet{perraudin2019deepspherecosmo}, the NN is made of $5$ graph convolutional layers, each followed by a \todo{max?} pooling layer down-sampling by $4$.
A GAP and a fully connected layer with softmax follow.
The Chebyshev polynomials are all of order $K=5$ and the number of channels per layer is $16, 32, 64, 64, 64$, respectively.
The cross-entropy loss is optimized with Adam for 80 epochs.
The learning rate is $2 \cdot 10^{-4} \cdot 0.999^{\textrm{step}}$ and the batch size is 8.

\begin{table}
    \centering
    \begin{tabular}{l c c c}
		\toprule
		& accuracy & conv. equiv. error & NN inv. error \\
		\midrule
		2D CNN baseline \citep{perraudin2019deepspherecosmo} & \todo{??} & \todo{can we?} \\
		8 neighbors, $\hat{t} = ??$, CNN variant, \citep{perraudin2019deepspherecosmo} & \todo{??} & \todo{??} \\
		8 neighbors, $\hat{t} = ??$, FCN variant, \citep{perraudin2019deepspherecosmo} & 83.8 & \todo{??} \\
		8 neighbors, optimal kernel width $\hat{t} = ??$ & 87.1 & \todo{??} \\
		20 neighbors, optimal kernel width $\hat{t} = ??$ & 91.25 & \todo{??} \\
		40 neighbors, optimal kernel width $\hat{t} = ??$ & \todo{??} & \todo{??} \\
		\bottomrule
    \end{tabular}
	\caption{
		Results on the classification of partial convergence maps.
		\todo{While extreme non-equivariance (the 2D CNN baseline) penalizes performance, improving equivariance doesn't improve performance.}
	}
    \label{tab:cosmo}
\end{table}

@@

The graph is built with \eqnref{weights:belkin}, \todo{$k = ?$} and the \todo{corresponding optimal kernel widths $t = ?, ?, ?$}.
Following \citet{jiang2019sphericalcnn}, the NN is an encoder-decoder with skip connections.
Details in \secref{climate:appendix}.
%The encoder is made of $5$ graph convolutional layers, each followed by a \todo{pick-one?} pooling layer down-sampling by $4$.
%the number of channels per layer is $16, 32, 64, 64, 64$, respectively.
%The decoder is made of
The Chebyshev polynomials are all of \todo{order $K=5$}.
The cross-entropy loss (weighted or non-weighted) is optimized with Adam for 30 epochs.
The learning rate is $1 \cdot 10{-3}$ and the batch size is 64.

@@

Results are shown in \tabref{climate} (details in tables~\ref{tab:climate:accuracy}, \ref{tab:climate:map} and~\ref{tab:climate:speed}).
The mean and standard deviation are computed over 5 runs.
Note that while \citet{jiang2019sphericalcnn} and \citet{cohen2019gauge} use a weighted cross-entropy loss, it is not ideal for the mAP metric.
Results from \citet{mudigonda2017climateevents} cannot be directly compared as they don't use the same input channels.
DeepSphere achieves state-of-the-art performance, suggesting again that anisotropic filters are unnecessary.

@@

\subsection{Uneven sampling} \label{sec:exp:ghcn}
% Weather prediction? But it's not a real task. Rather a demo.

To demonstrate the flexibility of modeling the discretized sphere by a graph, we collected historical measurements from about $n = 10,000$ weather stations scattered across the Earth.\footnote{\url{https://www.ncdc.noaa.gov/ghcn-daily-description}}
The spherical data is heavily non-uniformly sampled, with a much higher density of weather stations over North America than the Pacific (\figref{examples:ghcn:tmax}).
For illustration, we devised two artificial tasks.
A dense regression that consists of predicting the temperature on a given day knowing the temperature on the 5 previous days.
A global regression that consists of predicting the day (represented as one period of a sine over the year) from temperature or precipitations.

The graph is built with \eqnref{weights:khasanova} and \todo{$k = ?$} neighbors.
The kernel width $t$ is set as the average of the distances.
The equivariance property of the resulting graph has not been tested, and we don't expect it to be good due to the heavily non-uniform sampling.
The NN is made of $3$ graph convolutional layers.
The Chebyshev polynomials are all of order $K=0$ or $4$ and the number of channels per layer is $50, 100, 100$, respectively.
For the global regression, a GAP and a fully connected layer follow.
For the dense regression, a graph convolutional layer follows instead.
The MSE loss is optimized with RMSprop for 250 epochs.
The learning rate is $1 \cdot 10^{-3}$ and the batch size is 64.

\begin{table}
    \centering
    \begin{tabular}{
			c
			S[table-format=2.2]
			S[table-format=1.2]
			S[table-format=1.3]
			S[table-format=1.2]
			S[table-format=1.2]
			S[table-format=1.3]
			S[table-format=1.2]
			S[table-format=1.2]
			S[table-format=-1.3]
		}
		\toprule
		& \multicolumn{3}{c}{temp. (from past temp.)} & \multicolumn{3}{c}{day (from temperature)} & \multicolumn{3}{c}{day (from precipitations)} \\
		\cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(lr){8-10}
		order $K$ & {MSE} & {MAE} & {R2} & {MSE} & {MAE} & {R2} & {MSE} & {MAE} & {R2} \\
		\midrule
		$0$ & 10.88 & 2.42 & 0.896 & 0.10 & 0.10 & 0.881 & 0.58 & 0.42 & -0.920 \\
		%%$1$ &  8.91 & 2.20 & 0.906 & - & - & - & - & - & - \\
		$4$ &  8.20 & 2.11 & 0.919 & 0.05 & 0.06 & 0.969 & 0.50 & 0.18 &  0.597 \\
		%$9$ &  8.38 & 2.12 & 0.915 & - & - & - & - & - & - \\
		\bottomrule
    \end{tabular}
    \caption{
		Prediction results on data from weather stations.
		Structure always helps.
	}
    \label{tab:ghcn}
\end{table}

Results from \tabref{ghcn} show that treating the times series as a time-varying spherical signal (with filters of polynomial order $K>0$) yields higher performance than treating each time series independently ($K=0$).
Note that predicting the day from temperature is much easier as temperature has a yearly pattern.
Still, structure helps.

% * increasing order is good until threshold \\
% * equivariance to be verified in this case (next paper) -> in conclusion \\
% * The NN indeed predicts the correct temperature at day T, and not day T-1 (baseline) even though there is small difference. \\
% * \todo{performance of baseline} \\

>>>>

\todo{
\begin{itemize}
	\item shall we really name it? Or simply reference the general idea of a graph-based spherical CNN.
	\item adjacency matrix: $A$ instead of $W$?
	\item anonymize DeepSphere citation
% 	\item climate spherical figure: inputs with TC and AR labels superimposed [Frédérick]
	\item citet for inline, citep for parenthesis
	\item More details on related work? Not sure we have the space. If yes, detailed comparison with related work after method.
	\item PSD of climate data: how does it compare to cosmo and 3D objects?
	\item align the spheres in figure 1
	\item mention that \citep{esteves2018sphericalcnn} and \citep{cohen2018sphericalcnn} computed equivariance error
	\item check footnotes
	\item climate data on spectrum plot
	\item result tables: only what supports the main message in the text, everything we have in the appendix
	\item Unify vocabulary. Sampling, discretization, or pixelization of the sphere? Signal, maps, functions. Irregular, uneven. Coarsening, down-sampling.
	\item check English vs American English
	\item unify enumerations. I like (i), (ii), etc.
	\item notation: should we use indices from 1 to n instead of 0 to n-1 for compactness?
	\item add \citet{bruna2013gnn} in equivariance? They were the first to do a spherical CNN with graphs. Graph is built from a uniform random sampling of the sphere.
	\item figure placements
	\item do the experiments on the icosahedral grid with the pooling as defined in the paper
	\item put all the code (learning experiments from Frédérick and equivariance experiments from Martino) in a common reproducible git repository to be linked from the paper
	\item update pygsp and merge sphere branch
	\item check that NN architectures and hyper-params are correctly reported
	\item keep NN architecture and hyper-params in the main text? Move the paragraph to appendix if not enough space.
	%\item shall we still site adam and batch norm? no, mainstream enough
	\item check results for ModelNet40 and integrate in text
	\item Check that HEALPix is introduced in the introduction
	\item heuristic used to set the kernel width in v1. Add values to \figref{kernel_widths} for comparison.
	\item check experiments section [Frédérick]: all the results numbers and NN + training details
\end{itemize}
}
\newpage

@@

\paragraph{Architecture.}
All our NNs are fully convolutional, and employ a global average pooling (GAP) for rotation invariant tasks.
Graph convolutional layers are always followed by batch normalization and ReLU activation, except in the last layer.
Note that batch normalization and activation act on the elements of $\b{f}$ independently, and hence don't depend on the domain of $f$.
%We denote graph convolutional layers with $C$ output channels as $GC_C$ and fully connected layers with $C$ output neurons as $FC_C$.
%Those linear layers are always followed by batch normalization and ReLU activation, except in the last layer.
%softmax normalization as $SM$
%Pooling and unpooling by a factor $F$ are denoted as $P_F$ and $P^{-1}_F$, global average pooling as $GAP$.

@@

The recognition of 3D shapes is a rotation invariant task: rotating an object doesn't change its nature.
While 3D shapes are usually represented as meshes or point clouds, representing them as spherical maps (\figref{depthmap}) naturally allows a rotation invariant treatment.
%LIDAR is an example of a physical sensor that acquires spherical depth maps
%, which are a posteriori represented as point clouds.
% but there is gravity, might not really be SO(3) invariant
%\todo{move to intro?} no, good to motivate each task

@@

The equiangular graph is built with \eqnref{weights:khasanova} and $k = 4$ neighbors \citep[following][]{khasanova2017sphericalcnn}.
The HEALPix graph is built with \eqnref{weights:belkin}, $k = 8$, and a kernel width $t$ set to the average of the distances \citep[following][]{perraudin2019deepspherecosmo}.
\todo{Following Esteves or Cohen?}
The NN is made of $5$ graph convolutional layers, each followed by a max pooling layer down-sampling by $4$.
A GAP and a fully connected layer with softmax follow.
The Chebyshev polynomials are all of order $K=4$ and the number of channels per layer is $16, 32, 64, 128, 256$, respectively.
Following \citet{esteves2018sphericalcnn}, the cross-entropy plus a triplet loss is optimized with Adam for 30 epochs on the dataset augmented by 3 random translations.
The learning rate is $5 \cdot 10^{-2}$ and the batch size is 32.

@@

Results are shown in \tabref{shrec17}.
As the network is trained for shape classification rather than retrieval, we report the classification F1 score alongside the mAP used in the retrieval contest.\footnote{We omit the F1 for \citet{cohen2018sphericalcnn} as we didn't get the mAP reported in the paper when running it.}
DeepSphere achieves the same performance as \citet{cohen2018sphericalcnn} and \citet{esteves2018sphericalcnn} at a much lower cost, suggesting that anisotropic filters are an unnecessary price to pay.
% x times faster
As the information in those spherical maps resides in the low frequencies (\figref{spectrum}), reducing the equivariance error didn't translate into improved performance.
For the same reason, using the more uniform HEALPix discretization or lowering the resolution down to $N_{side} = 8$ didn't impact performance.
\todo{be sure that it fits with the cosmo conclusion}
%\todo{The same conclusions are drawn from ModelNet40 (\tabref{modelnet40}).}

@@

Graphs are built with \eqnref{weights:belkin}, $k = 8, 20, 40$ neighbors, and the corresponding optimal kernel widths $t$ given by the procedure in \secref{optimal}.
Following \citet{perraudin2019deepspherecosmo}, the NN is made of $5$ graph convolutional layers, each followed by a max pooling layer down-sampling by $4$.
A GAP and a fully connected layer with softmax follow.
The Chebyshev polynomials are all of order $K=5$ and the number of channels per layer is $16, 32, 64, 64, 64$, respectively.
The cross-entropy loss is optimized with Adam for 80 epochs.
The learning rate is $2 \cdot 10^{-4} \cdot 0.999^{\textrm{step}}$ and the batch size is 8.

\begin{table}
    \centering
    \begin{tabular}{l c c c}
		\toprule
		& accuracy & conv. equiv. error & NN inv. error \\
		\midrule
		2D CNN baseline \citep{perraudin2019deepspherecosmo} & \todo{??} & \todo{can we?} \\
		8 neighbors, $\hat{t} = ??$, CNN variant, \citep{perraudin2019deepspherecosmo} & \todo{??} & \todo{??} \\
		8 neighbors, $\hat{t} = ??$, FCN variant, \citep{perraudin2019deepspherecosmo} & 83.8 & \todo{??} \\
		8 neighbors, optimal kernel width $\hat{t} = ??$ & 87.1 & \todo{??} \\
		20 neighbors, optimal kernel width $\hat{t} = ??$ & 91.25 & \todo{??} \\
		40 neighbors, optimal kernel width $\hat{t} = ??$ & \todo{??} & \todo{??} \\
		\bottomrule
    \end{tabular}
	\caption{
		Results on the classification of partial convergence maps.
		\todo{While extreme non-equivariance (the 2D CNN baseline) penalizes performance, improving equivariance doesn't improve performance.}
		\todo{add speed to show trade-off}
	}
    \label{tab:cosmo}
\end{table}

@@

The graph is built with \eqnref{weights:belkin}, $k = 6$ neighbors, and a kernel width $t$ set to the average of the distances.
Following \citet{jiang2019sphericalcnn}, the NN is an encoder-decoder with skip connections.
Details in \secref{climate:appendix}.
%The encoder is made of $5$ graph convolutional layers, each followed by a \todo{pick-one?} pooling layer down-sampling by $4$.
%the number of channels per layer is $16, 32, 64, 64, 64$, respectively.
%The decoder is made of
The Chebyshev polynomials are all of order $K=4$.
The cross-entropy loss (weighted or non-weighted) is optimized with Adam for 30 epochs.
The learning rate is $1 \cdot 10{-3}$ and the batch size is 64.

@@

Results are shown in \tabref{climate} (details in tables~\ref{tab:climate:accuracy}, \ref{tab:climate:map} and~\ref{tab:climate:speed}).
The mean and standard deviation are computed over 5 runs.
Note that while \citet{jiang2019sphericalcnn} and \citet{cohen2019gauge} use a weighted cross-entropy loss, that is a suboptimal proxy for the mAP metric.
DeepSphere achieves state-of-the-art performance, suggesting again that anisotropic filters are unnecessary.
Note that results from \citet{mudigonda2017climateevents} cannot be directly compared as they don't use the same input channels.

@@

\subsection{Uneven sampling} \label{sec:exp:ghcn}
% time series modeling? GHCN?
% Weather prediction? But it's not a real task. Rather a demo.

To demonstrate the flexibility of modeling the discretized sphere by a graph, we collected historical measurements from $n \approx 10,000$ weather stations scattered across the Earth.\footnote{\url{https://www.ncdc.noaa.gov/ghcn-daily-description}}
The spherical data is heavily non-uniformly sampled, with a much higher density of weather stations over North America than the Pacific (\figref{examples:ghcn:tmax}).
For illustration, we devised two artificial tasks.
A dense regression: predict the temperature on a given day knowing the temperature on the previous 5 days.
A global regression: predict the day (represented as one period of a sine over the year) from temperature or precipitations.
Predicting from temperature is much easier as it has a clear yearly pattern.

The graph is built with \eqnref{weights:belkin}, $k = 5$ neighbors, and a kernel width $t$ set to the average of the distances.
The equivariance property of the resulting graph has not been tested, and we don't expect it to be good due to the heavily non-uniform sampling.
The NN is made of $3$ graph convolutional layers.
The Chebyshev polynomials are all of order $K=0$ or $4$ and the number of channels per layer is $50, 100, 100$, respectively.
For the global regression, a GAP and a fully connected layer follow.
For the dense regression, a graph convolutional layer follows instead.
The MSE loss is optimized with RMSprop for 250 epochs.
The learning rate is $1 \cdot 10^{-3}$ and the batch size is 64.

\begin{table}
    \centering
    \begin{tabular}{
			c
			S[table-format=2.2]
			S[table-format=1.2]
			S[table-format=1.3]
			S[table-format=1.2]
			S[table-format=1.2]
			S[table-format=1.3]
			S[table-format=1.2]
			S[table-format=1.2]
			S[table-format=-1.3]
		}
		\toprule
		& \multicolumn{3}{c}{temp. (from past temp.)} & \multicolumn{3}{c}{day (from temperature)} & \multicolumn{3}{c}{day (from precipitations)} \\
		\cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(lr){8-10}
		order $K$ & {MSE} & {MAE} & {R2} & {MSE} & {MAE} & {R2} & {MSE} & {MAE} & {R2} \\
		\midrule
		$0$ & 10.88 & 2.42 & 0.896 & 0.10 & 0.10 & 0.881 & 0.58 & 0.42 & -0.920 \\
		%%$1$ &  8.91 & 2.20 & 0.906 & - & - & - & - & - & - \\
		$4$ &  8.20 & 2.11 & 0.919 & 0.05 & 0.06 & 0.969 & 0.50 & 0.18 &  0.597 \\
		%$9$ &  8.38 & 2.12 & 0.915 & - & - & - & - & - & - \\
		\bottomrule
    \end{tabular}
    \caption{
		Prediction results on data from weather stations.
		Structure always improves performance.
	}
    \label{tab:ghcn}
\end{table}

Results are shown in \tabref{ghcn}.
While using a polynomial order $K=0$ is like modeling each time series independently with an MLP, orders $K>0$ integrate neighborhood information.
Results show that using the structure induced by the spherical geometry always yields better performance.

% * increasing order is good until threshold \\
% * The NN indeed predicts the correct temperature at day T, and not day T-1 (baseline) even though there is small difference. \\
% * \todo{performance of baseline} \\
